### 2022-12-09 intro.Rnw
new comments Leo

\begin{table}[!htb]
  \centering
  \caption{Statistical criteria for assessing replication success which have
    been used in practice \citep{Opensc2015, Camerer2016, Camerer2018, Cova2018,
      Errington2021}.}
  \label{tab:rsdef}
  \begin{tabular}{p{0.21\textwidth}p{0.7\textwidth}}
    \toprule
    Criterion type & Description \\
    \midrule
    Significance & The original and replication $p$-values are smaller than
                   a threshold $\alpha$ and their effect estimates show
                   the same direction
                   (\mbox{$p_{o} < \alpha$}, \mbox{$p_{r} < \alpha$}, and
                   \mbox{$\sign(\that_{o}) = \sign(\that_{r})$}). Usually $\alpha = 5\%$.
    \\
                   & \\
    Meta-analytic significance & The meta-analytic $p$-value is smaller than a threshold
                                 $\alpha$ (\mbox{$p_{m} = 2\{1 - \Phi(|\that_{m}|/\sigma_{m})\} < \alpha$})
                                 where \mbox{$\that_{m} =
                                 (\that_{o}/\sigma_{o}^{2} + \that_{r}/\sigma^{2}_{r})\sigma^{2}_{m}$}
                                 is the  pooled effect estimate with standard error
                                 \mbox{$\sigma_{m} = (1/\sigma^{2}_{o} + 1/\sigma^{2}_{r})^{-1/2}$}.
                                 Usually $\alpha = 5\%$. \\
                   & \\
    Relative effect size & The effect estimate of the replication study goes in the same
                           direction as the original one and is at least $d_{\text{min}}$ times as large (\mbox{$d = \that_{r}/\that_{o} \geq d_{\text{min}}$}).
                           Usually $d_{\text{min}} = 1$. \\
                   & \\
    Confidence interval & The replication effect estimate is contained in the
                          $(1 - \alpha)$ original confidence interval
                          ($\that_{r} \in [\that_{o} - z_{\alpha/2}\sigma_{o},
                          \that_{o} + z_{\alpha/2}\sigma_{o}]$).
                          Sometimes, also defined as the original effect estimate is
                          contained in the $(1 - \alpha)$ replication confidence interval
                          ($\that_{o} \in [\that_{r} - z_{\alpha/2}\sigma_{r},
                          \that_{r} + z_{\alpha/2}\sigma_{r}]$).
                          Usually $\alpha = 5\%$. \\
                   & \\
    Prediction interval ($Q$-test) & The replication effect estimate is contained in its
                                     $(1 - \alpha)$ prediction interval
                                     ($\that_{r} \in [\that_{o} - z_{\alpha/2}\surd(\sigma_{o}^{2}
                                     + \sigma_{r}^{2}), \that_{o} + z_{\alpha/2}\surd(\sigma_{o}^{2}
                                     + \sigma_{r}^{2})]$). Usually $\alpha = 5\%$.
                                     This criterion is equivalent to $p_{Q} \geq \alpha$ where
                                     $p_{Q}$ is the $p$-value from the meta-analytic $Q$-test
                                     $p_{Q} = 2\{1 - \Phi(\surd Q)\}$ with $Q$-statistic
                                     $Q = \sum_{i \in \{o, r\}} (\that_{i} - \that_{m})^{2}/\sigma_{i}^{2} = (\that_{o}- \that_{r})^{2}/(\sigma^{2}_{o} + \sigma^{2}_{r})$.\\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion is listed first and defines replication success by simultaneous
statistical significance of original and replication study along with their
effect estimates showing the same direction. This approach is also called the
``two-trials rule'' \citep{Senn2008} in drug regulation or ``vote-counting'' in
meta-analysis \citep{Cooper2019}. The criterion can similarly be defined through
one-sided $p$-values, so that the original and replication effect estimate are
not explicitly required to show the same direction as this is already taken into
account by the one-sided $p$-values. Some replication projects
\citep{Opensc2015, Errington2021} also defined replication success via
simultaneous non-significance of original and replication study
($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this definition
``replication success'' can almost always be achieved by conducting original and
replication study with very few samples so that the $p$-values are large. The
approach is also logically questionable as it could be seen as an instance of
the ``absence of evidence fallacy'' \citep{Altman1995} meaning the failure to
find evidence against the null hypothesis is erroneously interpreted as evidence
for the null hypothesis. Meta-analytic extensions of the significance approach
define replication success by significance of a combined effect
estimate.
% \footnote{The meta-analytic approach can also be given a Bayesian
% interpretation: A flat prior distribution for the underlying effect size
% $\theta$ is updated by the data from original and replication study.
% Replication success defined by a meta-analytic $p$-value being smaller than
% $\alpha$ is then equivalent to replication success via a Bayesian posterior
% probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
% respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
% depending on the direction of the combined estimate.}
Typically, the assumption of a common underlying effect size is seen as
reasonable so that fixed-effect meta-analysis is used for pooling.
Random-effects meta-analysis has primarily been used if more than one
replication study of the same original study are conducted since replicators are
often interested in understanding between-replication heterogeneity \citep[\eg{}
in][]{Klein2018}.

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the Reproducibility Project: Cancer Biology
    \citep{Errington2021}. (P, X, E) denotes effect number E from experiment
    number X, from original paper number P. Shown are original $\that_{o}$,
    replication $\that_{r}$, and pooled effect estimate $\that_{m}$ with 95\%
    confidence intervals, variance ratio $c = \sigma^{2}_{o}/\sigma^{2}_{r}$,
    relative effect estimate $d = \that_{r}/\that_{o}$, original
    \textit{p}-value $p_{o}$, replication \textit{p}-value $p_{r}$,
    meta-analytic \textit{p}-value $p_{m}$, $Q$-test \textit{p}-value $p_{Q}$,
    and sceptical \textit{p}-value $p_{\mathrm{S}}$. A replication success
    (green) threshold of $\alpha = 5\%$ is used for the \textit{p}-values and
    $d_{\text{min}} = 1$ for the relative effect estimate.}
  \label{tab:RPCB}
\resizebox{1.02\linewidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## function to format CI
ciFormat. <- function(lower, upper, dollar = FALSE, digits = 2) {
    if (dollar) x <- "$" else x <- ""
    paste0("[", x, format(round(lower, digits), nsmall = digits), x,
           " to ", x, format(round(upper, digits), nsmall = digits), x, "]")
}
ciFormat <- Vectorize(FUN = ciFormat.)

@@

## function to format (relative) effect estimate
dFormat. <- function(x, digits = 2, phantom = FALSE, dollar = FALSE) {
    if (is.na(x)) {
        out <- NA
    } else {
        out <- as.character(signif(x, digits = digits))
        if (phantom && (x >= 0)) {
           out <- paste0("\\phantom{-}", out)
        }
        if (dollar) out <- paste0("$", out, "$")
    }
    return(out)
}
dFormat <- Vectorize(FUN = dFormat.)

@@

## randomly select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(pr2) |>
    mutate(smdoFormat = signif(smdo, 3),
           smdrFormat = signif(smdr, 3),
           smdmFormat = signif(smdm, 3),
           dFormat = tableFormat(x = d, outFun = dFormat,
                                 threshold = 1, col1 = "dark2orange",
                                 col2 = "dark2green"),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,,
           po2Format = tableFormat(x = po2),
           pr2Format = tableFormat(x = pr2),
           pm2Format = tableFormat(x = pm2),
           ps2Format = tableFormat(x = ps2),
           pQFormat = tableFormat(x = pQ),
           cioFormat = ciFormat(lower = cioL, upper = cioU, dollar = TRUE),
           cirFormat = ciFormat(lower = cirL, upper = cirU, dollar = TRUE),
           cimFormat = ciFormat(lower = cimL, upper = cimU), dollar = TRUE) |>
    select(id,
           smdoFormat, cioFormat,
           smdrFormat, cirFormat,
           smdmFormat, cimFormat,
           c, dFormat,
           po2Format, pr2Format, pm2Format, pQFormat, ps2Format
           )
rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$", "[95\\% CI]",
                          "$\\that_r$", "[95\\% CI]",
                          "$\\that_m$", "[95\\% CI]",
                          "$c$",
                          "$d$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )

@@

To illustrate reverse-Bayes inference, let us return to the replication setting.
Assume we want to conduct inference about the unknown effect size $\theta$ based
on the effect estimate from the original study $\that_{o}$ and its standard
error $\sigma_{o}$. We will assume that $\that_{o}$ is normally distributed
around the unknown effect size $\theta$ with (known) variance equal to its
squared standard error $\sigma^{2}_{o}$, here and henceforth denoted by
$\that_{o} \given \theta \sim \Nor(\theta, \sigma^{2}_{o})$. Furthermore, we
specify a zero-mean normal prior with variance $\tau^{2}$ for the effect size
$\theta$, representing the position of a sceptic who does not believe in the
presence of a genuine effect. The ``stubbornness'' of the sceptic is determined
by how small the variance $\tau^{2}$ is chosen (the smaller $\tau^{2}$, the more
scepticism). Combining the sceptical prior with the likelihood produces a
posterior which is again normal
$\theta \given \that_{o}, \sigma_{o} \sim \Nor(\mu_{\text{post}}, \sigma^{2}_{\text{post}})$
with mean and variance
\begin{align*}
  &\mu_{\text{post}} = \frac{\that_{o}}{1 + \sigma^{2}_{o}/\tau^{2}}&
  &\text{and}&
  &\sigma^{2}_{\text{post}} = \frac{1}{1/\sigma^{2}_{o} + 1/\tau^{2}}.&
\end{align*}
The associated $(1- \alpha)$ highest posterior density credible interval has
limits
\begin{align}
  \left[\mu_{\text{post}} - z_{\alpha/2} \, \sigma_{\text{post}},
  \mu_{\text{post}} + z_{\alpha/2} \, \sigma_{\text{post}}\right]
  \label{eq:postCI}
\end{align}
and if this credible interval excludes parameter values smaller/larger than zero
(depending on the orientation of the effect size) this may be interpreted as
evidence for a genuine effect found in the original study. There exist also
other definitions of statistical evidence and in subsequent chapters of this
thesis (Paper\ref{PaperII} and Paper\ref{PaperIV}) we will extend this
reverse-Bayes procedure to employ alternative measures of evidence, such as
Bayes factors.

@@

This thesis consists of six papers divided into three parts. The first part
(Paper~\ref{PaperI}, \ref{PaperII}, and \ref{PaperIII}) focuses on methodology
for design and analysis of replication studies. The second part
(Paper~\ref{PaperIV} and \ref{PaperV}) revolves around reverse-Bayes methodology
with a broader scope of applications than replication
studies. % Paper IV is a review paper about
% reverse-Bayes methodology and Paper V is a short comment on another article
% which proposed a reverse-Bayes method.
The last part (Paper~\ref{PaperVI}) is on meta-scientific perspectives on
methodological research.
% The article lists and illustrates questionable
% research practices in simulation studies, and gives recommendations how to avoid
% them.

>>>>

\begin{table}[!htb]
  \centering
  \caption{Statistical criteria for assessing replication success which have
    been used in practice \citep{Opensc2015, Camerer2016, Camerer2018, Cova2018,
      Errington2021}.}
  \label{tab:rsdef}
  \begin{tabular}{p{0.21\textwidth}p{0.7\textwidth}}
    \toprule
    Criterion type & Description \\
    \midrule
    Significance & The original and replication $p$-values are smaller than
                   a threshold $\alpha$ and their effect estimates show
                   the same direction
                   (\mbox{$p_{o} < \alpha$}, \mbox{$p_{r} < \alpha$}, and
                   \mbox{$\sign(\that_{o}) = \sign(\that_{r})$}). Usually $\alpha = 5\%$.
    \\
                   & \\
    Meta-analytic significance & The meta-analytic $p$-value is smaller than a threshold
                                 $\alpha$ (\mbox{$p_{m} = 2\{1 - \Phi(|\that_{m}|/\sigma_{m})\} < \alpha$})
                                 where \mbox{$\that_{m} =
                                 (\that_{o}/\sigma_{o}^{2} + \that_{r}/\sigma^{2}_{r})\sigma^{2}_{m}$}
                                 is the  pooled effect estimate with standard error
                                 \mbox{$\sigma_{m} = (1/\sigma^{2}_{o} + 1/\sigma^{2}_{r})^{-1/2}$}.
                                 There are no established conventions regarding the level $\alpha$.
    \\
                   & \\
    Relative effect size & The effect estimate of the replication study goes in the same
                           direction as the original one and is at least $d_{\text{min}}$ times as large (\mbox{$d = \that_{r}/\that_{o} \geq d_{\text{min}}$}).
                           Usually $d_{\text{min}} = 1$. \\
                   & \\
    Confidence interval & The replication effect estimate is contained in the
                          $(1 - \alpha)$ original confidence interval
                          ($\that_{r} \in [\that_{o} - z_{\alpha/2}\sigma_{o},
                          \that_{o} + z_{\alpha/2}\sigma_{o}]$).
                          Sometimes, also defined as the original effect estimate is
                          contained in the $(1 - \alpha)$ replication confidence interval
                          ($\that_{o} \in [\that_{r} - z_{\alpha/2}\sigma_{r},
                          \that_{r} + z_{\alpha/2}\sigma_{r}]$).
                          Usually $\alpha = 5\%$. \\
                   & \\
    Prediction interval ($Q$-test) & The replication effect estimate is contained in its
                                     $(1 - \alpha)$ prediction interval
                                     ($\that_{r} \in [\that_{o} - z_{\alpha/2}\surd(\sigma_{o}^{2}
                                     + \sigma_{r}^{2}), \that_{o} + z_{\alpha/2}\surd(\sigma_{o}^{2}
                                     + \sigma_{r}^{2})]$). Usually $\alpha = 5\%$.
                                     This criterion is equivalent to $p_{Q} \geq \alpha$ where
                                     $p_{Q}$ is the $p$-value from the meta-analytic $Q$-test
                                     $p_{Q} = 2\{1 - \Phi(\surd Q)\}$ with $Q$-statistic
                                     $Q = \sum_{i \in \{o, r\}} (\that_{i} - \that_{m})^{2}/\sigma_{i}^{2} = (\that_{o}- \that_{r})^{2}/(\sigma^{2}_{o} + \sigma^{2}_{r})$.\\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:rsdef} lists commonly used criteria for replication success in
terms of the summary statistics from Table~\ref{tab:meta}. The most popular
criterion is listed first and defines replication success by simultaneous
statistical significance of original and replication study along with their
effect estimates showing the same direction. This approach is also called the
``two-trials rule'' \citep{Senn2008} in drug regulation or ``vote-counting'' in
meta-analysis \citep{Cooper2019}. The criterion can similarly be defined through
one-sided $p$-values, so that the original and replication effect estimate are
not explicitly required to show the same direction as this is already taken into
account by the one-sided $p$-values. Some replication projects
\citep{Opensc2015, Errington2021} also defined replication success via
simultaneous non-significance of original and replication study
($p_{o} > \alpha$ and $p_{r} > \alpha$). However, with this definition
``replication success'' can almost always be achieved by conducting original and
replication study with very few samples so that the $p$-values are large. The
approach is also logically questionable as it could be seen as an instance of
the ``absence of evidence fallacy'' \citep{Altman1995} meaning the failure to
find evidence against the null hypothesis is erroneously interpreted as evidence
for the null hypothesis. Meta-analytic extensions of the significance approach
define replication success by significance of a combined effect
estimate.
% \footnote{The meta-analytic approach can also be given a Bayesian
% interpretation: A flat prior distribution for the underlying effect size
% $\theta$ is updated by the data from original and replication study.
% Replication success defined by a meta-analytic $p$-value being smaller than
% $\alpha$ is then equivalent to replication success via a Bayesian posterior
% probability $\Pr(\theta > 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
% respectively, $\Pr(\theta < 0 \given \that_{o}, \that_{r}) > 1 - \alpha/2$,
% depending on the direction of the combined estimate.}
Typically, the assumption of a common underlying effect size is seen as
reasonable so that fixed-effect meta-analysis is used for pooling.
Random-effects meta-analysis has primarily been used if more than one
replication study of the same original study are conducted since replicators are
often interested in understanding between-replication heterogeneity \citep[\eg{}
in][]{Klein2018}. In contrast to the ordinary significance criterion, it is less
established which level $\alpha$ should be used for thresholding the
meta-analytic $p$-value. Some replication projects have used the same level as
for the significance approach \citep{Opensc2015, Camerer2016, Errington2021},
whereas others have used a smaller level as the pooled estimate is based on both
data sets \citep[\eg{} $\alpha = 0.5\%$ in][]{Camerer2018}.

@@

\begin{landscape}
\begin{table}
  \centering
  \caption{Subset of results from the Reproducibility Project: Cancer Biology
    \citep{Errington2021}. (P, X, E) denotes effect number E from experiment
    number X, from original paper number P. Shown are original $\that_{o}$,
    replication $\that_{r}$, and pooled effect estimate $\that_{m}$ with 95\%
    confidence intervals, variance ratio $c = \sigma^{2}_{o}/\sigma^{2}_{r}$,
    relative effect estimate $d = \that_{r}/\that_{o}$, original
    \textit{p}-value $p_{o}$, replication \textit{p}-value $p_{r}$,
    meta-analytic \textit{p}-value $p_{m}$, $Q$-test \textit{p}-value $p_{Q}$,
    and sceptical \textit{p}-value $p_{\mathrm{S}}$. A replication success
    (green) threshold of $\alpha = 5\%$ is used for the \textit{p}-values and
    $d_{\text{min}} = 1$ for the relative effect estimate.}
  \label{tab:RPCB}
\resizebox{1.02\linewidth}{!}{%
<< "RPCB-table2", results = "asis" >>=
## function to format CI
ciFormat. <- function(lower, upper, dollar = FALSE, digits = 2) {
    if (dollar) x <- "$" else x <- ""
    paste0("(", x, format(round(lower, digits), nsmall = digits), x,
           " to ", x, format(round(upper, digits), nsmall = digits), x, ")")
}
ciFormat <- Vectorize(FUN = ciFormat.)

@@

## function to format (relative) effect estimate
dFormat. <- function(x, digits = 2, phantom = FALSE, dollar = FALSE) {
    if (is.na(x)) {
        out <- NA
    } else {
        out <- format(round(x, digits), nsmall = digits)
        if (phantom && (x >= 0)) {
           out <- paste0("\\phantom{-}", out)
        }
        if (dollar) out <- paste0("$", out, "$")
    }
    return(out)
}
dFormat <- Vectorize(FUN = dFormat.)

@@

## randomly select a subset with one effect per paper
set.seed(7)
rpcbTable <- rpcb |>
    na.omit() |>
    group_by(paper) |>
    sample_n(size = 1) |>
    ungroup() |>
    arrange(pr2) |>
    mutate(smdoFormat = signif(smdo, 3),
           smdrFormat = signif(smdr, 3),
           smdmFormat = signif(smdm, 3),
           dFormat = tableFormat(x = d, outFun = dFormat,
                                 threshold = 1, col1 = "dark2orange",
                                 col2 = "dark2green"),
           cioL = smdo - za*so,
           cioU = smdo + za*so,
           cirL = smdr - za*sr,
           cirU = smdr + za*sr,
           cimL = smdm - za*sm,
           cimU = smdm + za*sm,,
           po2Format = tableFormat(x = po2),
           pr2Format = tableFormat(x = pr2),
           pm2Format = tableFormat(x = pm2),
           ps2Format = tableFormat(x = ps2),
           pQFormat = tableFormat(x = pQ),
           cioFormat = ciFormat(lower = cioL, upper = cioU, dollar = TRUE),
           cirFormat = ciFormat(lower = cirL, upper = cirU, dollar = TRUE),
           cimFormat = ciFormat(lower = cimL, upper = cimU), dollar = TRUE) |>
    select(id,
           smdoFormat, cioFormat,
           smdrFormat, cirFormat,
           smdmFormat, cimFormat,
           c, dFormat,
           po2Format, pr2Format, pm2Format, pQFormat, ps2Format
           )
rpcbXtable <- xtable(rpcbTable)
colnames(rpcbXtable) <- c("(P, X, E)",
                          "$\\that_o$", "(95\\% CI)",
                          "$\\that_r$", "(95\\% CI)",
                          "$\\that_m$", "(95\\% CI)",
                          "$c$",
                          "$d$",
                          "$p_o$",
                          "$p_r$",
                          "$p_m$",
                          "$p_Q$",
                          "$p_{\\text{S}}$"
                          )

@@

To illustrate reverse-Bayes inference, let us return to the replication setting.
Assume we want to conduct inference about the unknown effect size $\theta$ based
on the effect estimate from the original study $\that_{o}$ and its standard
error $\sigma_{o}$. We will assume that $\that_{o}$ is normally distributed
around the unknown effect size $\theta$ with (known) variance equal to its
squared standard error $\sigma^{2}_{o}$, here and henceforth denoted by
$\that_{o} \given \theta \sim \Nor(\theta, \sigma^{2}_{o})$. Furthermore, we
specify a zero-mean normal prior with variance $\tau^{2}$ for the effect size
$\theta$, representing the position of a sceptic who does not believe in the
presence of a genuine effect. The ``stubbornness'' of the sceptic is determined
by how small the variance $\tau^{2}$ is chosen (the smaller $\tau^{2}$, the more
scepticism). Combining the sceptical prior with the likelihood produces a
posterior which is again normal
$\theta \given \that_{o}, \sigma_{o} \sim \Nor(\mu_{\text{post}}, \sigma^{2}_{\text{post}})$
with mean and variance
\begin{align*}
  &\mu_{\text{post}} = \frac{\that_{o}}{1 + \sigma^{2}_{o}/\tau^{2}}&
  &\text{and}&
  &\sigma^{2}_{\text{post}} = \frac{1}{1/\sigma^{2}_{o} + 1/\tau^{2}}.&
\end{align*}
The associated $(1- \alpha)$ highest posterior density credible interval has
limits
\begin{align}
  \mu_{\text{post}} \pm z_{\alpha/2} \, \sigma_{\text{post}}
  \label{eq:postCI}
\end{align}
and if this credible interval excludes parameter values smaller/larger than zero
(depending on the orientation of the effect size) this may be interpreted as
evidence for a genuine effect found in the original study. There exist also
other definitions of statistical evidence and in subsequent chapters of this
thesis (Paper~\ref{PaperII} and Paper~\ref{PaperIV}) we will extend this
reverse-Bayes procedure to employ alternative measures of evidence, such as
Bayes factors.

@@

This thesis consists of six papers divided into three parts. The first part
(Paper~\ref{PaperI}, \ref{PaperII}, and \ref{PaperIII}) focuses on methodology
for design and analysis of replication studies. The second part
(Paper~\ref{PaperIV} and \ref{PaperV}) revolves around reverse-Bayes methodology
with a broader scope of applications than replication
studies. % Paper IV is a review paper about
% reverse-Bayes methodology and Paper V is a short comment on another article
% which proposed a reverse-Bayes method.
The last part (Paper~\ref{PaperVI}) deals with meta-scientific perspectives of
methodological research.
% The article lists and illustrates questionable
% research practices in simulation studies, and gives recommendations how to avoid
% them.
